---
title: "Chap 1<br>Principal Component Analysis"
author: "François Lefebvre"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  prettydoc::html_pretty:
      toc: yes
      toc_depth: 2
      number_sections: true
      theme: architect
      highlight: github
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(paletteer)
library(patchwork)
library(here)
library(FactoMineR)
library(factoextra)
library(simstudy)

knitr::opts_chunk$set(echo = TRUE,
                      results = "hide")
```



```{r include=FALSE}
theme_set(
  ggthemes::theme_tufte(base_size = 11, base_family = "sans", ticks = TRUE) +
  theme(title = element_text(color = "midnightblue"),
        panel.grid.major  = element_blank(),
        panel.grid.minor  = element_blank(),
        plot.background = element_rect(fill = "gainsboro", color = NA))
  )
```




# Data - Notation - Examples

Let $\bar{x}_k$ be the mean of variable $k$ calculated over all individual
instances of $I$.

$$
\bar{x}_k = \frac{1}{I}\sum_{i=1}^I{x_{ik}}
$$

and the standard deviation is

$$
s_k = \sqrt{\frac{1}{I} \sum_{i=1}^I{(x_{ik}-\bar{x}_k})^2}
$$


# Objectives

The orange orange data

```{r}
orange <- list()
orange$data <- read.csv(here("orange.csv")) %>%
  mutate(Brand = factor(Brand))
# data used for factorial PCA
orange$num <- orange$data %>%
  select(where(is.numeric))
skimr::skim(orange$data)
```
## Studying Individuals

```{r}
sim <- list()
sim <- within(sim, {
  defs <- defData(varname = "group", dist = "categorical", formula = "0.5;0.5")
  defs <- defData(defs, varname = "j", dist = "normal", formula = "ifelse(group%%2, -1, 1)", variance = 0.1)
  defs <- defData(defs, varname = "k", dist = "normal", formula = "ifelse(group%%2, -1, 1)", variance = 0.1)
  data <- genData(defs, n = 40)
})
p <- list()
p$p1 <- ggplot(sim$data, aes(j, k, color = abs(j * k))) +
  geom_point(show.legend = FALSE) +
  scale_color_paletteer_c("oompaBase::bluescale") +
  coord_fixed()
sim <- within(sim, {
  defs <- defData(varname = "j", dist = "uniform", formula = "-1;1")
  defs <- defData(defs, varname = "k", dist = "normal", formula = "j", variance = 0.05)
  data <- genData(defs, n = 40)
})
p$p2 <- ggplot(sim$data, aes(j, k, color = abs(j * k))) +
  geom_point(show.legend = FALSE) +
  scale_color_paletteer_c("oompaBase::greenscale") +
  coord_fixed()
# p$p2
sim <- within(sim, {
  defs <- defData(varname = "j", dist = "normal", formula = 0, variance = 0.5)
  defs <- defData(defs, varname = "k", dist = "normal", formula = 0, variance = 0.5)
  data <- genData(defs, n = 40)
})
p$p3 <- ggplot(sim$data, aes(j, k, color = abs(j * k))) +
  geom_point(show.legend = FALSE) +
  scale_color_paletteer_c("oompaBase::redscale") +
  coord_fixed()
# p$p3
wrap_plots(p)
```

### Example - orange



```{r}
orange$pca <- orange$num %>%
  PCA(scale.unit = TRUE, graph = FALSE)
fviz_pca_var(orange$pca) | fviz_pca_ind(orange$pca)
```

```{r}
condes(orange$num, num.var = 2)
```


```{r}
dimdesc(orange$pca)
```

# Studying Individuals


# Studying Variables


# Relationship between $N_I$ and $N_K$


# Interpreting the Data


# Implementation with `FactoMineR`



# Additional Results



# Example: Decathlon



